---
title: "US Weekly Gas Price Volatility"
subtitle: "Proposal"
author: 
  - name: "Elizabeth Ladd"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project will look at a dataset containing weekly gas prices in the U.S. to determin how common sudden price increases are and if there is a difference between regular or diesel gas."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(dplyr)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
weekly_gas_prices <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-07-01/weekly_gas_prices.csv')
# code source: https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-07-01/readme.md

glimpse(weekly_gas_prices) #gets overview of data

```

This dataset was taken from the TidyTuesday github repo from 2025-07-01. It looks at the weekly US gas prices for both regular and diesel. The date range of this data set is from Aug 1990 to June 2025. I mainly chose this data set because I wanted to explore some of the peaks of previous spikes in gas prices, especially with the volatility of gas prices so far in 2026.

From the TidyTuesday README: "The data comes from the U.S. Energy Information Administration (EIA), which publishes average retail gasoline and diesel prices each Monday." The original data can be found at [Gasoline and Diesel Fuel Update - U.S. Energy Information Administration (EIA)](https://www.eia.gov/petroleum/gasdiesel/ "https://www.eia.gov/petroleum/gasdiesel/")

## Questions

Are sudden spikes in gas prices common, or do prices rise at a relatively constant rate?

Are there differences in volatility of prices between regular and diesel?

## Analysis plan

- For the first question, I will just look at regular gas prices. Also, to make it consistent over time, I will look at "regular" grade and "all" formulation. I will add two new variables to calculate the change in price from the previous week and the percent change in order to visualize trends in "spikes."
-  For the second question, I will need the percent change as in the first question, but I am more concerned with the movement in either direction, so I will use the absolute value of the change to measure volatility. I will need to compare the "fuel" variable, selecting "gasoline" and "diesel"

| Task | Status | Due | Priority | Summary |
|----|----|----|----|----|
| Review TidyTuesday data sets and select one | Complete | 31 May | Medium |  |
| Complete Project Proposal for Peer Review | Complete | 4 June | High |  |
| Submit Revised Proposal | Pending peer review | 9 June | Medium |  |
| Develop Question 1 answers | Ongoing | 14 June | Medium |  |
| Develop Question 2 answers | Ongoing | 21 June | Medium |  |
| Write final write up | Pending presentation completion | 26 June | Medium |  |
| Record presentation | Pending presentation completeion | 28 June | Medium |  |
| Submit final write up and presentation | Pending peer review | 30 June | High |  |
